---
title: "Sample Size Calculation"
author: "Prof A Pathmeswaran<br>pathmes@kln.ac.lk<br>2022-08-30"
format: 
  revealjs:
   theme: "simple"
   embed-resources: true
   standalone: true
---

## Outline

1.  Background
2.  Basics
3.  Common situations

## Why do it?

1.  Researcher
    -   Ensure useful results
    -   Plan logistics
2.  Higher degree committees
    -   Assess feasibility
    -   Usefulness of results
3.  Funding agency
    -   Value for money

## Based on the research question

-   Research questions/ objective
    1.  Estimate a mean or percentage with a given precision - descriptive study.
    2.  Identify the association between two variables/ difference between two groups - testing a hypothesis or analytical study. Difference between two means, the difference between two percentages, a correlation coefficient, etc.

## Descriptive study

-   Estimating population value based on sample findings
-   The larger the sample size, the more precise the estimate (narrower confidence interval)
-   The larger the sample size, the higher the confidence

\

-   The higher the variability, the larger the sample size
    -   Numeric variable - standard deviation\
    -   Categorical variable - % closer to 50%

## Descriptive study - Numeric outcome

-   Required information
    1.  Acceptable difference (d) half the width of the 95% Confidence Interval
    2.  An estimate of the standard deviation (sd)

N is the required sample size

$$N = \frac{1.96^2 * sd^2}{d^2}$$

## Descriptive study - Categorical outcome

-   Required information
    1.  Acceptable difference (d) half the width of the 95% Confidence Interval
    2.  An estimate of the expected percentage (p)

N is the required sample size

$$N = \frac{1.96^2 * p(100-p)}{d^2}$$

## Analytical Study

-   Avoid type I error (declaring as significant by mistake)
    -   p \< 0.05
    -   p \< 0.01
-   Ensure adequate power (detecting an association if it is there)
    -   80%
    -   90%

## Analytical study - Numeric outcome {.smaller}

-   Required information
    1.  Expected difference/ effect size (d)
    2.  Estimate of the two standard deviations (sd~1~ and sd~2~)

*n* is the required sample size **per group** (significance 0.05 & power 80%).

$$n = \frac{(1.96 + 0.84)^2(sd_1^2 + sd_2^2)}{d^2}$$

If significance 0.01 & power 90%) $$n = \frac{(2.58 + 1.28)^2(sd_1^2 + sd_2^2)}{d^2}$$

## Analytical study - Categorical outcome

-   Required information
    1.  Expected difference/ effect size (p~1~ - p~2~)
    2.  Estimate of the two percentages (p~1~ and p~2~)

*n* is the required sample size **per group** (significance 0.05 & power 80%).

$$n = \frac{(1.96 + 0.84)^2[(p_1(100- p_1) + p_2(100-p_2)]}{(p_1 - p_2)^2}$$

## Sampling methods and sample size

-   Modify calculation based on the sampling method
    -   Cluster samples result in larger standard errors
    -   Unequal group sizes increase the total sample size

## What if the sample size is fixed?

-   Secondary data
-   Feasibility
-   Cost

## Fixed sample size

-   Indicate what is possible with that sample size
    -   Expected confidence interval for a percentage
-   Example
    -   The available sample size is 1000
    -   Expected 95% CI if the % is 1, 2.5, 5, 10, 25, 40 or 50

## Expected confidence intervals

```{r}
#| message: false
#| warning: false
#| echo: false
library(dplyr)
library(Hmisc)
library(gt)
x <- c(10, 25, 50, 100, 250, 400, 500)
n <- 1000
a <- binconf(x, n)
c <- a*100
ct <- as_tibble(c)

gt(ct) %>% 
  fmt_number(
    columns = 2:3,
        decimals = 1
  ) %>% 
as_raw_html()

```

## Fixed sample size

-   Analytical study
    -   Calculate expected power for different scenarios
-   Example
    -   The available sample size is 1000
    -   An equal number of exposed & unexposed
    -   \% with outcome
    -   10 & 15
    -   10 & 20

## Expected power

| Size1 | Prev1 | Size2 | Prev2 | Power |
|-------|-------|-------|-------|-------|
| 500   | 10    | 500   | 15    | 63%   |
| 500   | 10    | 500   | 20    | 96%   |
| 900   | 10    | 100   | 20    | 81%   |

[OpenEpi](https://www.openepi.com/Menu/OE_Menu.htm)

## Power and sample size

(Sig = 0.05, P~1~ = 0.15 & p~2~ = 0.1)

```{r}
#| message: false
#| warning: false
#| echo: false

library(pwr)
p1 <- .15
p2 <- .1
h= 2*asin(sqrt(p1))-2*asin(sqrt(p2))
p.2p <- pwr.2p.test(h=h,power = 0.8, sig.level=0.05,alternative="greater")
plot(p.2p)

```

## Power and sample size

(Sig = 0.01, P~1~ = 0.15 & p~2~ = 0.1)

```{r}
#| message: false
#| warning: false
#| echo: false

library(pwr)
p1 <- .15
p2 <- .1
h= 2*asin(sqrt(p1))-2*asin(sqrt(p2))
p.2p <- pwr.2p.test(h=h,power = 0.8, sig.level=0.01,alternative="greater")
plot(p.2p)

```

## Power and sample size

(P~1~ = 0.5 & p~2~ = 0.42)

```{r}
#| message: false
#| warning: false
#| echo: false

library(pwr)
p1 <- .5
p2 <- .42
h= 2*asin(sqrt(p1))-2*asin(sqrt(p2))
p.2p <- pwr.2p.test(h=h,power = 0.8, sig.level=0.05,alternative="greater")
plot(p.2p)

```

## Recap

-   Background
    -   Study designs
    -   Standard error and confidence intervals
    -   Significance and power
-   Basic Sample size calculations
-   Fixed sample size

[Home](https://pathmes.quarto.pub/sample-size-calculation/)
